---
title: "EDA"
format: html
editor: visual
---

# Exploratory Data Analysis of the Diabetes Health Indicators Dataset

##### Annie DiFrank

##### ST558 Final Project, December 5, 2024

## Introduction

In this project, we will be exploring the Diabetes Health Indicators Dataset (available on [Kaggle](https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset/)).

Diabetes is a serious chronic disease with multiple complications associated. There are many genetic, environmental, and lifestyle factors that can be associated with risk of diabetes. The risk of type II diabetes, the most common form, can differ by race, education, age, income, and many additional factors.

The data analyzed here was collected via a health-related telephone survey conducted by the CDC's Behavioral Risk Factor Surveillance System (BRFSS). These features are either questions directly answered by participants or calculated variables based on participants' responses. We are interested in finding the best model to predict the occurance of prediabetes or diabetes (a binary variable, Diabetes_binary). That is, we are interested in what risk factors are the most predictive of diabetes, and can we use this subset of variables to classify disease occurrence?

There are already well-known medical factors (mainly related to diet) that can increase risk of diabetes, so for this project I am going to focus on some factors that may have less-known associations; mental health, sex, age, education, income, and smoker status.

Especially because we did not personally create this data, the purpose of this exploratory analysis is to get to know the data and how it is stored. This includes overall distributions of variables, how variables are stored (numerically, as factor variables, etc- do the column types make sense for the variable) and any pre-existing relationships between variables that we should be aware of before attempting to fit the data to a predictive model.

With a well-fit model, we could potentially create a risk stratification panel for individuals that have not been diagnosed with prediabetes/diabetes.

In this file, we will perform exploratory data analysis before moving on to modeling with the data.

#### Variable Information

| Variable        | Data Type       | Information                                                                                |
|-------------|-------------|----------------------------------------------|
| Diabetes_binary | binary          | 0 = no diabetes 1 = prediabetes or diabetes                                                |
| Smoker          | binary          | Have you smoked at least 100 cigarettes in your entire life? 0 = no 1 = yes                |
| MentHlth        | integer/numeric | For how many days during the past 30 days was your mental health not good? scale 1-30 days |
| Sex             | binary          | 0 = female 1 = male                                                                        |
| Age             | integer         | 13-level age category (\_AGEG5YR see codebook) 1 = 18-24 9 = 60-64 13 = 80 or older        |
| Education       | integer/factor  | Education level (EDUCA see codebook) scale 1-6                                             |
| Income          | integer/factor  | Income scale (INCOME2 see codebook)                                                        |

: Click [here](https://www.cdc.gov/brfss/annual_data/2014/pdf/CODEBOOK14_LLCP.pdf) for the codebook!

```{r setup, include=FALSE}
#pacman to check whether packages are installed, if not load them
if (!require("pacman")) install.packages("pacman")
library(pacman)
pacman::p_load(dplyr,
 tidyverse,
 purrr,
 ggplot2,
 readr,
 psych,
 reshape2,
 GGally)

```

## Data Import

```{r dataimport}
 diabetes <- read_csv("app/diabetes_binary_health_indicators_BRFSS2015.csv")
```

## Exploratory Analysis/ Clean up

```{r cleanup}
str(diabetes)
# all variables have been imported as numeric, going to adjust appropriately- 

diabetes <- diabetes %>% 
  mutate(Diabetes_binary = factor(Diabetes_binary, levels = c(0, 1), labels = c("No", "Yes")),
         Smoker = factor(Smoker, levels = c(0, 1), labels = c("No", "Yes")),
        Sex = factor(Sex, levels = c(0, 1), labels = c("Female", "Male")),
        Age = factor(Age, levels = c(1,2,3,4,5,6,7,8,9,10,11,12,13), labels = c("18-24","25-29","30-34","35-39","40-44","45-49","50-54","55-59","60-64","65-69","70-74","75-79","80 or older")),
        Education = factor(Education, levels = c(1,2,3,4,5,6), labels = c("never attended school or only kindergarten","elementary","some high school","high school graduate","some college or technical school","college graduate")),
        Income = factor(Income, levels = c(1,2,3,4,5,6,7,8), labels = c("less than 10k","less than 15k", "less than 20k", "less than 25k","less than 35k","less than 50k","less than 75k","75 or more"))
         ) %>% 
  select(c(Diabetes_binary, Smoker, Sex, Age, Education, Income, MentHlth))

#Checking for missingness 
sum(is.na(diabetes))
  #no NAs 

describe(diabetes)
 #note, all the categorical variables have * next to the variable name.


```

## Data Summarizations

#### First, let's look at how our 6 variables of interest interact by themselves with diabetes/pre-diabetes occurrence.

```{r}
diabetes %>% group_by(Diabetes_binary) %>% summarize(count=n())

```

```{r smokerstatus}
diabetes %>% group_by(Diabetes_binary,Smoker) %>% summarize(count=n())

ggplot(data = diabetes, aes(x=Smoker, fill=Diabetes_binary)) +
  geom_bar() +
  labs(x="Smoker Status") +
  scale_fill_discrete("Diabetes")
```

There doesn't appear to be any notable association here.

```{r agestatus}
diabetes %>% group_by(Diabetes_binary,Age) %>% summarize(count=n())

ggplot(data = diabetes, aes(x=Age, fill=Diabetes_binary)) +
  geom_bar() +
  labs(x="Age, 5 year bins") +
  scale_fill_discrete("Diabetes")
```

I expected diabetes occurance to have the positive trend with age. It visually looks like it goes down at 70+ years old, but I believe this is just due to the total N of these age groups being lower.

```{r sexstatus}
diabetes %>% group_by(Diabetes_binary,Sex) %>% summarize(count=n())

ggplot(data = diabetes, aes(x=Sex, fill=Diabetes_binary)) +
  geom_bar() +
  labs(x="Sex") +
  scale_fill_discrete("Diabetes")
```

No big difference here.

```{r Incomestatus}
diabetes %>% group_by(Diabetes_binary,Income) %>% summarize(count=n())

ggplot(data = diabetes, aes(x=Income, fill=Diabetes_binary)) +
  geom_bar() +
  labs(x="Income") +
  scale_fill_discrete("Diabetes")
```

One thing you could glean here is that there are over double the amount of people that make more than 75k versus less than 75k (and every other wealth status for that matter) but it doesn't appear that diabetes occurance doubles- looks like income could potentially be a protective factor to a certain extent. This would make sense; better access to healthier foods, health care, etc.

```{r Educationstatus}
diabetes %>% group_by(Diabetes_binary,Education) %>% summarize(count=n())

ggplot(data = diabetes, aes(x=Education, fill=Diabetes_binary)) +
  geom_bar() +
  labs(x="Education") +
  scale_fill_discrete("Diabetes")
```

No notable associatons here. I wish they were able to collect a more equal amount of people across all education statuses.

```{r menthlthstatus}
diabetes %>% group_by(Diabetes_binary) %>% summarize(mean_MentHlth=mean(MentHlth),median_MentHlth=median(MentHlth))

ggplot(data = diabetes) +
  geom_boxplot(aes(x = Diabetes_binary, y=MentHlth, fill=Diabetes_binary)) +
  labs(y="Mental Health, Rated 1-30 Days Feeling Poorly")

```

It looks like those with diabetes have experienced more poor mental health days than those without, which makes sense. It would be interesting to see that variable further split up- what caused that poor mental health (job? physical health? relationship stress?)

#### Next, lets look at any variable correlations. I'm just going to explore some relationships I think could be significant.

```{r}
ggplot(data=diabetes) +
  geom_boxplot(aes(x=Income, y=MentHlth, fill = Income))
```

It looks like the median of poor mental health days decreases as income levels increase. Lower-income groups have a higher spread/variability than higher-income groups. In general, higher income appears to correlate with fewer days of poor mental health on average.

```{r}
ggplot(data=diabetes) +
  geom_boxplot(aes(x=Education, y=MentHlth, fill = Education))
```

We see similar trends here; it appears that the median amount of poor days decreases as education levels increase.

```{r}
mosaicplot(table(diabetes$Income, diabetes$Education), 
           color = TRUE, 
           main = "Mosaic Plot of Income and Education",
           xlab = "Income", 
           ylab = "Education")
```

Education and income appear to have a positive correlation.

```{r}
mosaicplot(table(diabetes$Smoker, diabetes$Sex), 
           color = TRUE, 
           main = "Mosaic Plot of Smoker status and Sex",
           xlab = "Smoker", 
           ylab = "Sex")
```

In this population if you are a male you are more likely to be a smoker than if you are a female.

```{r}
ggplot(data = diabetes, aes(x=Sex, fill =Smoker)) +
  geom_bar(position = "dodge") +
  labs(x="Sex") +
  scale_fill_discrete("Smoker") +
  facet_wrap(~ Education)
  
```

Here we see that for lower education levels (elementary, some highschool, highschool graduate, some college) there are more male smokers than female (relative to the total count in each group- there is a higher percentage of men that smoke that dont) but in the highest education level, for both sexes it is less common to be a smoker.

```{r}
ggplot(data = diabetes, aes(x=Sex, fill =Smoker)) +
  geom_bar(position = "dodge") +
  labs(x="Sex") +
  scale_fill_discrete("Smoker") +
  facet_wrap(~ Income)
```

We see the same trends here- again highlighting the association between income and education it appears.

## Conclusions

The biggest thing we have gathered from this analysis is that there is an association between two predictor variables, education and income. There are likely other positive associations that just make logical sense (like age and smoking, or age and education or income level). We also saw some potential associations between mental health and other predictor variables as well as with diabetes occurrence.

Next, lets model. Click [here](https://anniedifrank168.github.io/ST558_finalproject/Modeling.html) to go to the Modeling section
