---
title: "Modeling"
format: html
editor: visual
---

# Modeling Analysis of the Diabetes Health Indicators Dataset

##### Annie DiFrank

##### ST558 Final Project, December 5, 2024

## Introduction

```{r setup, include=FALSE}
#pacman to check whether packages are installed, if not load them
if (!require("pacman")) install.packages("pacman")
library(pacman)
pacman::p_load(tidycensus,
 dplyr,
 tidyverse,
 httr,
 jsonlite,
 purrr,
 ggplot2,
 readr,
 reshape2,
 gganimate)

```

## Data Import
```{r dataimport}
 diabetes <- read_csv("diabetes_binary_health_indicators_BRFSS2015.csv")
```

## Splitting the Data 
```{r datasplit}

```

## Classification Tree Modeling

You should provide a thorough explanation of what a classification tree model is. Then you should fit a classification tree with varying values for the complexity parameter and choose the best model (based on 5 fold CV on the training set). Include at least 5 predictors in this model.
```{r classtree}

```

## Random Forest Modeling

You should provide a thorough explanation of what a random forest is and why we might use it (be sure to relate this to a basic classification tree). You should then fit a random forest model with varying values for the mtry parameter and choose the best model (based on 5 fold CV on the training set). Include at least 5 predictors in this model.

```{r rftree}

```

## Final Model Selection
```{r}

```

